A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-09 latex/main.tex
Updates from Overleaf

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
depth and richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore such single-tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.

>>>>

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore simple tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
These are the Type I events, in which the ARG reflects the
hierarchical structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.
